Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors

### 2013-10-14 Direction_JINST_v3.tex
Direction_JINST_v1.tex has the same content as Direction_v7.tex but is in JINST format. The references have been adjusted to match the JINST journal style. We use JHEP.bst as recommended in the JINST author manual. Direction_JINST_v2.tex has an additional section with three additional plots containing results for 1.4 and 2.1 MeV. Direction_JINST_v3.tex is another iteration of Direction_JINST_v2.tex and Direction_JINST_v4.tex is the version which Andrey, Henry, Matt and Christoph propose for publication. Some more details can be found in README.md --Christoph

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.5]{graphs/PEandR_vs_energy.pdf}
        \caption[]{The energy dependence of the mean number of PEs after the 34.0~ns time cut is shown for Cherenkov-induced PEs (black circles) and scintillation-induced PEs (black squares). The ratio between the mean number of Cherenkov-induced and scintillation-induced PEs is shown as blue triangles, values are given on the right y-axis. The statistical errors are too small to be seen. \label{Edep_NPE}}
        \end{center}
\end{figure}

@@

\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hDir_pq_1p4.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hVtx_pq_1p4.pdf}}
        \caption[]{(Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons
        (1.4~MeV). In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        from the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation. \label{low_energy_reco}}
\end{center}
\end{figure}
\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs/hCos_vs_E.pdf}
        \caption[]{Mean cosine of the angle between the true initial electron direction and the reconstructed direction, depending on the energy of the initial electrons. For each energy 1000 events have been simulated. Statistical errors are
shown. \label{Edep_angle}}
        \end{center}
\end{figure}

>>>>

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.5]{graphs/PEandR_vs_energy_line.pdf}
        \caption[]{The energy dependence of the mean number of PEs after the 34.0~ns time cut is shown for Cherenkov-induced PEs (black circles) and scintillation-induced PEs (black squares). The ratio between the mean number of Cherenkov-induced and scintillation-induced PEs is shown as blue triangles connected by a line, values are given on the right y-axis. The statistical errors are too small to be seen. \label{Edep_NPE}}
        \end{center}
\end{figure}

@@

\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hDir_pq_1p4.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hVtx_pq_1p4.pdf}}
        \caption[]{(Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons
        (1.4~MeV). In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        from the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation. \label{low_energy_reco}}
\end{center}
\end{figure}
\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs/hCos_vs_E.pdf}
        \caption[]{Mean cosine of the angle between the true initial electron direction and the reconstructed direction, as a function of the electron energy. For each energy 1000 events have been simulated. Statistical errors are
shown. \label{Edep_angle}}
        \end{center}
\end{figure}

### 2013-10-14 Direction_JINST_v4.tex
Direction_JINST_v1.tex has the same content as Direction_v7.tex but is in JINST format. The references have been adjusted to match the JINST journal style. We use JHEP.bst as recommended in the JINST author manual. Direction_JINST_v2.tex has an additional section with three additional plots containing results for 1.4 and 2.1 MeV. Direction_JINST_v3.tex is another iteration of Direction_JINST_v2.tex and Direction_JINST_v4.tex is the version which Andrey, Henry, Matt and Christoph propose for publication. Some more details can be found in README.md --Christoph

\documentclass[cits]{JINST}
%%\usepackage{multirow}
\usepackage{times}
%%\usepackage{hyperref}
\usepackage{lineno}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{mathptmx}
\usepackage{mdwlist}
\usepackage{textcomp}
\usepackage{graphicx,subfigure}
\usepackage{fontenc}
%%\usepackage{color}

@@

\keywords{Scintillators, scintillation and light emission processes (solid, gas and liquid scintillators); Cherenkov detectors; Large detector systems for particle and astroparticle physics; Neutrino detectors; Simulation methods and programs}

@@

The studies in sections \ref{sim_section} to \ref{reconstruction_sec} are made with single 5~MeV electrons, lower energies are
discussed in section \ref{edep_section}. Future studies will work on increasing the complexity
of the event as is needed for $0\nu\beta\beta$ studies. Three
effects primarily contribute to the timing of the scintillator detector
system. First, the simulated travel time of the initial 5~MeV electron is typically between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5m-radius detector at the level of precision needed for
direction reconstruction.

@@

\section{Energy dependence}
\label{edep_section}
In the previous sections we presented results on single 5~MeV electrons.
In this section we study two lower energies, 1.4~MeV and 2.1~MeV. These
energies correspond to Q/2 for the double beta decay of
$^{116}$Cd and $^{48}$Ca, respectively\cite{cd1,biller}. The isotope
$^{116}$Cd was chosen because of its potential use in quantum-dot-doped
scintillators\cite{qdot,qdot2} and $^{48}$Ca was chosen
to cover the Q-value range of $0\nu\beta\beta$ candidate isotopes.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.5]{graphs/PEandR_vs_energy_line.pdf}
        \caption[]{The energy dependence of the mean number of PEs after the 34.0~ns time cut is shown for Cherenkov-induced PEs (black circles) and scintillation-induced PEs (black squares). The ratio between the mean number of Cherenkov-induced and scintillation-induced PEs is shown as blue triangles connected by a line, values are given on the right y-axis. The statistical errors are too small to be seen. \label{Edep_NPE}}
        \end{center}
\end{figure}

@@

\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hDir_pq_1p4.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hVtx_pq_1p4.pdf}}
        \caption[]{(Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons
        (1.4~MeV). In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        from the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation. \label{low_energy_reco}}
\end{center}
\end{figure}
\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs/hCos_vs_E.pdf}
        \caption[]{Mean cosine of the angle between the true initial electron direction and the reconstructed direction, as a function of the electron energy. For each energy 1000 events have been simulated. Statistical errors are
shown. \label{Edep_angle}}
        \end{center}
\end{figure}

@@

\acknowledgments
The authors thank Andrew Blake at University of Cambridge for his work
authoring the WCSimAnalysis code. The authors thank the neutrino
reconstruction group at Iowa State, particularly Mayly Sanchez, Ioana
Anghel, and Tian Xin, for their continued work in developing the
WCSimAnalaysis algorithms and for their insights and expertise
regarding issues related to Cherenkov reconstruction with
fast-timing. The authors also thank Micheal Smy for his development of
the quadruplet-based vertex-finding method. L. Winslow would like to
thank Janet Conrad for many useful discussion on the topic, and
Katsushi Arisaka for discussions on the possible reach of traditional PMTs
and the characteristics of HPDs. C. Aberle and L. Winslow are
supported by funds from University of California Los Angeles. The work
at the University of Chicago is partially supported by DOE
contract DE-SC0008172 and NSF grant PHY-1066014. Matthew Wetstein gratefully
acknowledges support by the Grainger Foundation.

>>>>

\pdfoutput=1
\documentclass[cits]{JINST}

%%\usepackage{multirow}
\usepackage{times}
%%\usepackage{hyperref}
\usepackage{lineno}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{mathptmx}
\usepackage{mdwlist}
\usepackage{textcomp}
\usepackage{graphicx,subfigure}
\usepackage{fontenc}
%%\usepackage{color}

@@

\keywords{Scintillators, Large detector systems for particle and astroparticle physics; Neutrino detectors; Simulation methods and programs}

@@

The studies in sections \ref{sim_section} to \ref{reconstruction_sec} are made with single 5~MeV electrons, lower energies are
discussed in section \ref{edep_section}. Future studies will work on increasing the complexity
of the event as is needed for $0\nu\beta\beta$ studies. Three
effects primarily contribute to the timing of the scintillator detector
system. First, the simulated travel time of an initial 5~MeV electron is typically between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5m-radius detector at the level of precision needed for
direction reconstruction.

@@

\section{Energy dependence}
\label{edep_section}
In the previous sections we presented results on single 5~MeV electrons.
In this section we study two lower energies, 1.4~MeV and 2.1~MeV. These
energies correspond to Q/2 for the double-beta decay of
$^{116}$Cd and $^{48}$Ca, respectively\cite{cd1,biller}. The isotope
$^{116}$Cd was chosen because of its potential use in quantum-dot-doped
scintillators\cite{qdot,qdot2} and $^{48}$Ca was chosen
to cover the Q-value range of $0\nu\beta\beta$ candidate isotopes.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.5]{graphs/PEandR_vs_energy_final.pdf}
        \caption[]{The energy dependence of the mean number of PEs after the 34.0~ns time cut is shown for Cherenkov-induced PEs (black open circles, dotted line) and scintillation-induced PEs (black open squares, dashed line). The ratio between the mean number of Cherenkov-induced and scintillation-induced PEs is shown as blue filled triangles, values are given on the right y-axis. The statistical errors are too small to be seen. \label{Edep_NPE}}
        \end{center}
\end{figure}

@@

\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hDir_pq_1p4.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hVtx_pq_1p4.pdf}}
        \caption[]{(Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons
        (1.4~MeV). In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        from the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation. \label{low_energy_reco}}
\end{center}
\end{figure}
\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs/hCos_vs_E_final.pdf}
        \caption[]{Mean cosine of the angle between the true initial electron direction and the reconstructed direction, as a function of the electron energy. For each energy 1000 events have been simulated. Statistical errors are
shown. \label{Edep_angle}}
        \end{center}
\end{figure}

@@

\acknowledgments
The authors thank Andrew Blake at University of Cambridge for his work
authoring the WCSimAnalysis code. The authors thank the neutrino
reconstruction group at Iowa State, particularly Mayly Sanchez, Ioana
Anghel, and Tian Xin, for their continued work in developing the
WCSimAnalysis algorithms and for their insights and expertise
regarding issues related to Cherenkov reconstruction with
fast-timing. The authors also thank Michael Smy for his development of
the quadruplet-based vertex-finding method. L. Winslow would like to
thank Janet Conrad for many useful discussion on the topic, and
Katsushi Arisaka for discussions on the possible reach of traditional PMTs
and the characteristics of HPDs. C. Aberle and L. Winslow are
supported by funds from University of California Los Angeles. The work
at the University of Chicago is partially supported by DOE
contract DE-SC0008172 and NSF grant PHY-1066014. Matthew Wetstein gratefully
acknowledges support by the Grainger Foundation.
